---
title: "Installs.Rmd"
author: "April Wright"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installs

Workshop participants should have a working laptop with R installed. RStudio is strongly suggested. 

You can download the whole workshop website and lessons from [here](https://github.com/darunabas/biogeography-workshop).

Necessary R packages:

```{r eval = FALSE, echo=TRUE}

install.packages("phyloregion")
install.packages("ape")
install.packages("phytools")
install.packages("phangorn") 
install.packages("tidyverse") 
install.packages("devtools")
install.packages("pkgdown")
install.packages("knitr")
install.packages("rmarkdown")

```

If you choose not to use RStudio, a text editor that can edit plain text and Markdown files is also necessary. BBEdit or TextEdit are fine. 

Notes:

Day1: 12/11
- "ape" package
- How to use "vignette" function
> vignette(package="phangorn")
> vignette("Trees",package="phangorn")

- a phylogenetic tree
```{r phylogenetic_tree, eval = FALSE, echo=TRUE}
library(ape)
library(phangorn)
tree <- read.tree(text = "(, (, ));")
plot(tree, no.margin = TRUE)
nodelabels()
tiplabels()
# to understand how the trees are ordered --> ?reorder.phylo()
str(tree)
tree$edge

random <- rtree(5) # generate a random phylogeny with a specified number of tips
ladder_random <- ladderize(random) # reorder the tree to get a better visual
plot(ladder_random, no.margin=TRUE)
nodelabels()
tiplabels()
ladder_random$edge

post_tree <- reorder(ladder_random, "postorder")
plot(post_tree, no.margin=TRUE)
nodelabels()
tiplabels()
ladder_random$edge


getNumDes <- function(df_edge, num_mom){
    num_daughters <- nrow(df_edge[df_edge$mom == num_mom,])
    return(num_daughters)
}

getLiDes <- function(tree_obj){
    df_edge <- as.data.frame(tree_obj$edge)
    colnames(df_edge) <- c('mom', 'daughter')
    li_mom <- unique(df_edge$mom)
    
    df_output <- as.data.frame(li_mom)
    colnames(df_output) <- c('mom')
    df_output$num_des <- 0
    
    for(i in li_mom){
        df_output[df_output$mom == i, ]$num_des <- getNumDes(df_edge, i)
    }
    
    return(df_output)
}


nr_desc <-  function(x){
    x <- reorder(x, "postorder") # doesn't affect plotting but the edge matrix
    node_count <- numeric(max(x$edge)) # initialize
    node_count[1:Ntip(x)] = 1L # set the tips to have a descendant of 1
    # why filling tips with 1 is really important
    for(i in 1:nrow(x$edge)){
        tmp = x$edge[i,1]
        node_count[tmp] =  node_count[tmp] + node_count[x$edge[i,2]]
    }
    node_count
}

nr_desc(tree)
df_tree_count <- getLiDes(tree) # this one does not count all but good try to myself!

# solution shared by Jennie
function1 <- function(x){
    x <- x$edge
    y <- x[~which(x[,2]==nrow(x)==1),] # to exclude the root
    table(y[,1])
}

# solution by Dr. Dave
treeParts <- prop.part(tree)
#sapply simplifies output from lapply
sapply(treeParts, length)
nKids <- function(tree){
    sapply(prop.part(tree), length)
}
nKids(tree) # spits out number of kids except for the tips

```

```{r multiphylo,eval = FALSE, echo=TRUE}
trees <- rmtree(1000, 100) # random multi tree, 1000 trees each with 100 tips
trees # 1000 phylogenetic trees
trees[[1]]
sapply(trees, Ntip)

read.tree()

write.tree(trees, "many_trees.tre")
write.nexus(trees, "many_trees.nex")


```